Sampbias, a method to evaluate geographic sampling bias in species distribution data

### 2019-12-03 manuscript.rmd
Update manuscript.rmd

Once we have an estimated value of $l$, we can infer the expected accessibility bias as a function of distance using $b(x,l)$. Since the function $b(x,l)$ describes the exponentially decreasing sampling probability in relation to distance, we can define a standardized bias function as: $B(x,l) = 1 - b(x,l)/b(0,l)$, 
[[*Actually I think this is effectively equivalent to what I worteabove $b(x,l) = \exp(-lx)$*]]
where  the level of bias is set to 0 at distance 0 from the bias source. The standardized bias function tends asymptotically to 1 as the distance tends to infinity. However, since in any area $max(x) << \infty$, for small values of $l$ (i.e. little or no bias), $B(x,l)$ will look essentially like a uniform distribution with values very close to 0. Large values of $l$ (i.e. strong bias) will instead result in a curve that quickly approaches 1 with increasing distance. The values provided by the standardized bias function can be interpreted as the proportion of occurrences that are missing from the sample, compared to the observed samples at distance 0. Thus, if for a given estimated $l$ we have $B(50,l) = 0.20$, we can expect that at 50 Km from e.g. a road the number of occurrences per grid cell will be about 80% of the occurrences sampled at distance 0 from the road, with 20% missing due to sampling bias.

When running *sampbias*, we typically test different sources of biases, such as roads, cities, airports, etc. For each factor an independent expected distribution ($f(x)$) is computed and a parameter $l$ is estimated from $g(x,l)$. These estimates can be use to produce maps showing the intensity of potential biases across the area based on the standardized bias function. The bias values obtained from different sources can then be averaged in each grid cell to produce a map showing the combined effects of all sources.

>>>>

Once we have an estimated value of $l$, we can infer the expected accessibility bias as a function of distance using $b(x,l)$. Since the function $b(x,l)$ describes the exponentially decreasing sampling probability in relation to distance, we can define a standardized bias function as: $B(x,l) = 1 - b(x,l)/b(0,l)$, 
[[*Actually I think this is effectively equivalent to what I wrote above $b(x,l) = \exp(-lx)$*]]
where  the level of bias is set to 0 at distance 0 from the bias source. The standardized bias function tends asymptotically to 1 as the distance tends to infinity. However, since in any area $max(x) << \infty$, for small values of $l$ (i.e. little or no bias), $B(x,l)$ will look essentially like a uniform distribution with values very close to 0. Large values of $l$ (i.e. strong bias) will instead result in a curve that quickly approaches 1 with increasing distance. The values provided by the standardized bias function can be interpreted as the proportion of occurrences that are missing from the sample, compared to the observed samples at distance 0. Thus, if for a given estimated $l$ we have $B(50,l) = 0.20$, we can expect that at 50 Km from e.g. a road the number of occurrences per grid cell will be about 80% of the occurrences sampled at distance 0 from the road, with 20% missing due to sampling bias.

When running *sampbias*, we typically test different sources of biases, such as roads, cities, airports, etc. For each factor an independent expected distribution ($f(x)$) is computed and a parameter $l$ is estimated from $g(x,l)$. These estimates can be use to produce maps showing the intensity of potential biases across the area based on the standardized bias function. The bias values obtained from different sources can then be averaged in each grid cell to produce a map showing the combined effects of all sources.
[[*yeah this part is still weird. Basically, if rivers are not explaining anything they should count nothing toward the combined estimate, whereas they do now. I think we could try to see if the max likelihoods are comparable and weight the average by that.*]]
